---
title: "Data Tidying"
author: "Mike Loranty"
date: "1/16/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning = F, message = F}
library(dplyr)
library(tidyr)
```

To call a function from a specific package type 'package_name::function_name()'

# Data Tidying

Read in the data file

```{r}
catch_df <- read.csv(file = "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1",
                  stringsAsFactors = FALSE)

# for Windows there could be a problem, if so, use the following
#read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", 
#               method = "libcurl"),stringsAsFactors = FALSE)

head(catch_df)
```

cmd + shift + m is the keyboard shortcut for the pipe operator %>%

This section will make the catch data frame tidy

*  Remove marginal sum and notes column 
*  Move from wide to long format 

```{r}
catch_long <- catch_df %>%
  select(-All, -notesRegCode) %>% 
  # in gather you need to specify columns - this can be positive or negative
  gather(key = "species", value = "catch", -Year, -Region)

head(catch_long)
```

* erroneous value due to OCR (Optical Character Recogniton) issues  
* create catch column in correct units

```{r}
catch_cleaned <- catch_long %>% 
  rename(catch_thousands = catch) %>% 
  mutate(catch_thousands = ifelse(catch_thousands == "I",1,catch_thousands)) %>% 
  mutate(catch_thousands = as.integer(catch_thousands)) %>% 
  mutate(catch = catch_thousands*1000)

tail(catch_cleaned)

```


```{r, eval = F, echo = F}
# we used this code to find the bad value 
test_catch <- as.integer(catch_cleaned$catch_thousands)

i <- which(is.na(test_catch)==T)

catch_cleaned[i,]
```

# Split-Apply-Combine

Calculate total catch of all species by region

```{r}
catch_total <- catch_cleaned %>% 
  group_by(species, Year) %>% 
  summarize(catch_mean = mean(catch),
            catch_sd = sd(catch),
            n_obs = n()) # this n() function gives the number of observations in each

catch_total

```

Filter the data set to include only data from chinook salmon

```{r}
catch_chinook <- catch_cleaned %>% 
  filter(species == "Chinook" | species == "Chum") %>% 
  arrange(Year) # use a (-) negative to sort in descending order 

head(catch_chinook)
```

# Spread

Make a wide dataframe using the spread function

```{r}
catch_wide <- catch_cleaned %>% 
  filter(Year > 1990) %>% 
  select(-catch_thousands) %>% 
  spread(key = Year, value = catch)

head(catch_wide)
```


# Joins


```{r}
region_defs <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1",
                            method = "libcurl"),stringsAsFactors = FALSE)

head(region_defs)
```

Clean up region definitions data
```{r}
region_clean <- region_defs %>% 
  select(code,mgmtArea) 

head(region_clean)
  
```

```{r}

catch_joined <- left_join(catch_cleaned,region_clean,
                          by = c("Region" = "code"))

head(catch_joined)
```

# Separate and Unite

```{r}
dates_df <- data.frame(date = c("5/24/1930",
                                "5/25/1930",
                                "5/26/1930",
                                "5/27/1930",
                                "5/28/1930"),
                       stringsAsFactors = FALSE)

dates_sep <- dates_df %>% 
  separate(date, c("month", "day", "year"), sep = "/",
           remove = F) # this keep the original column

dates_sep
```


```{r}
dates_unite <- dates_sep %>% 
  unite(date_iso, year, month, day, sep = "-")

dates_unite
```


