---
title: "data_visualization"
author: "Mike Loranty"
date: "1/17/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning = F, message = F}
library(dplyr)
library(tidyr)
library(ggplot2)
library(DT)
```

# Data Tidying

Read in data from Alaska Department of Fish and Game, Division of Commercial Fisheries. 2017. Daily salmon escapement counts from the OceanAK database, Alaska, 1921-2017. Knowledge Network for Biocomplexity. doi:10.5063/F1S46Q6M.

```{r}
esc <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e",
                stringsAsFactors = F)

#head(esc) comment this out, or include 'echo = F, results = F' above
```

Calculate annual, regional, total escapement  

* `separate` the data into year - month - day  
* `group_by` `summarize` to calculate sum over region-year-species combinations  
* `filter` for salmon species  

```{r}
annual_esc <- esc %>% 
  separate(sampleDate, into = c("year", "month", "day"), sep = "-") %>% 
  mutate(year = as.numeric(year)) %>% 
  group_by(SASAP.Region, Species, year) %>% 
  summarize(escapement = sum(DailyCount)) %>% 
  filter(Species %in% c("Sockeye", "Chinook", "Pink", "Chum", "Coho")) 
```

```{r}
datatable(annual_esc)
```

# Static Plots

```{r}
# use the plus sign to add functions, etc... into ggplot
# everything in aes call should be mapping an asepct of the plot to a column in the dataset
ggplot(data = annual_esc, mapping = aes(x = Species, y = escapement, fill = SASAP.Region)) +
  geom_col() 
```

Timeseries of Kodiak escapement by species

```{r}
# you can create a theme object with settings that can be added to a ggplot object
# my_theme <- theme_bw() +
#   theme(plot.title = element_text(hjust = 0.5))
# this is good for many plots

ggplot(data = filter(annual_esc, SASAP.Region == "Kodiak"), 
       aes(x = year, y = escapement, color = Species)) + 
  geom_line() + 
  geom_point() + 
  ylab(expression(paste(alpha, " Escapement (number of fish)"^2))) + 
  ggtitle(expression(paste("Kodiak Salmon Escpament ", delta, N^15))) + 
  # can use bquote() for this stuff too
  theme_bw()
  #theme(plot.title = element_text(hjust = 0.5)) # this aligns the title in the center
# or + my_theme


```


```{r, fig.height = 10, fig.width = 6}
p <- ggplot(annual_esc, aes(x = year, y = escapement, color = Species)) +
  geom_line() + 
  geom_point() +
  facet_wrap(~SASAP.Region, scales = "free_y", ncol = 2)
# ggsave("region_escapement.png", plot = p)

# can also use grid.arrange to make multipanel plots with different types of plots/data
```


# Interactive Map
